---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data##


- Data source


[Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)


```{r, echo = TRUE}

library(utils)
unzip("activity.zip")
activity <- read.csv("activity.csv")

##saving date-time from date and interval

date <- as.Date(activity$date)
time <- sprintf("%04d", activity$interval)
datetime <- paste(date,time, sep =" ")
datetime <- strptime(datetime, format = "%Y-%m-%d %H%M", tz ="")
activity$datetime <- datetime

head(activity)
```

 

## What is mean total number of steps taken per day?
```{r, echo = TRUE}

#sum up the steps taken by day into a new vector

aggregatedata <- aggregate(activity$steps, by = list(date), FUN = sum)
names(aggregatedata)<- c("date", "avgstepsday")
head(aggregatedata)

mean(aggregatedata$avgstepsday, na.rm = TRUE)
median(aggregatedata$avgstepsday, na.rm = TRUE)

```



```{r, echo = TRUE}

hist(aggregatedata$avgstepsday, col = "grey", xlab = "No. of steps per day", main = "No. of steps per day")
abline(v = mean(aggregatedata$avgstepsday, na.rm = TRUE), col = "blue")
text(13000, 25, "mean = 10,770", col = "blue")
rug(aggregatedata$avgstepsday)


```



## What is the average daily activity pattern?

```{r, echo = TRUE}
aggbyinterval <- aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm = TRUE)
names(aggbyinterval)<- c("interval", "avgsteps")
head(aggbyinterval)

```

**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
```{r, echo = TRUE}
aggbyinterval[which.max(aggbyinterval$avgsteps),]$interval
```

```{r, echo = TRUE}
with(aggbyinterval, plot(interval, avgsteps, type = "l", main = "Average Daily Activity", xaxt = "n"))
axis(1, at = aggbyinterval$interval, cex.axis = 0.75)

abline(v = aggbyinterval[which.max(aggbyinterval$avgsteps),]$interval, col = "blue", lwd = 2)
text(aggbyinterval[which.max(aggbyinterval$avgsteps),]$interval + 200, 200, "max at 0835 = 206.17", col = "blue")
```




## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, echo = TRUE}
sum(is.na(activity$steps))

```


**2304 rows with missing values**

Devise a strategy for filling in all of the missing values in the dataset. 

Replacing missing values with the avg. steps taken during that time interval across all days

```{r, echo = TRUE}
#merging avg. steps by interval into a new data set
newactivity <- merge(activity, aggbyinterval, by = "interval")
newactivity <- newactivity[order(newactivity$date),]
head(newactivity)

newactivity$steps[is.na(newactivity$steps)] <- newactivity$avgsteps[is.na(newactivity$steps)]
head(newactivity)

```


Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r, echo = TRUE}
newagg <- aggregate(newactivity$steps, by = list(date), FUN = sum)
names(newagg) <- c("date","avgstepsday")
summary(newagg$avgstepsday)

hist(newagg$avgstepsday, col = "grey", xlab = "No. of steps per day", main = "No. of steps per day")
abline(v = mean(newagg$avgstepsday, na.rm = TRUE), col = "blue")
rug(newagg$avgstepsday)

```

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

**The median is closer to the mean once imputed values are used**

## Are there differences in activity patterns between weekdays and weekends?


```{r, echo=TRUE}

newactivity$weekend <- weekdays(newactivity$datetime, abbreviate = TRUE) == "Sun" | weekdays(newactivity$datetime, abbreviate = TRUE) == "Sat"

newactivity$weekday <- !newactivity$weekend

## long way to get avg steps by intervals across weekends and weekdays separately. Still learning how to create functions so this is all a bit manual. 

weekdaydata <- subset(newactivity, weekend == FALSE)
weekenddata <- subset(newactivity, weekend == TRUE)

aggbyintervalWeekend <- aggregate(weekenddata$steps, by = list(weekenddata$interval), FUN = mean, na.rm = TRUE)
names(aggbyintervalWeekend)<- c("interval", "avgsteps2")
weekenddata <- merge(weekenddata, aggbyintervalWeekend, by = "interval")

aggbyintervalWeekday <- aggregate(weekdaydata$steps, by = list(weekdaydata$interval), FUN = mean, na.rm = TRUE)
names(aggbyintervalWeekday)<- c("interval", "avgsteps2")
weekdaydata <- merge(weekdaydata, aggbyintervalWeekday, by = "interval")

library(gridExtra)
require(gridExtra)
library(lattice)

plot1<- xyplot(avgsteps2 ~ interval, data = weekenddata, type ="l", ylab ="avg. no of steps", main ="weekend")

plot2<- xyplot(avgsteps2 ~ interval, data = weekdaydata, type ="l", ylab = "avg. no of steps", main ="weekday")

grid.arrange(plot1,plot2, nrow = 2)

```

